Inference in Hidden Markov Models with Explicit State Duration Distributions

### 2011-10-27 tex/edhmm.tex
updated paths
small edits, mostly to do with computation complexity. One graph left...

\documentclass[journal]{oldtran/IEEEtran}
\usepackage{amsmath,amssymb,graphicx,subfig}
\usepackage[usenames,dvipsnames]{color}
\usepackage{algorithm,algorithmic}
\newcommand{\mike}[1]{\textcolor{red}{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}}}
\newcommand{\frank}[1]{\textcolor{red}{\textsf{\emph{\textbf{\textcolor{blue}{#1}}}}}}
\newcommand{\simeon}[1]{\textcolor{red}{\textsf{\emph{\textbf{\textcolor{green}{#1}}}}}}
\newcommand{\E}{\textrm{E}}

\title{Inference in Hidden Markov Models with Explicit State Duration Distributions}
\author{\mike{Mike}, Chris, \frank{Frank}}

@@

    \begin{figure}[t]
        \centering
        \subfloat[][]{
            \includegraphics[width=0.25\textwidth]{EDHMM_graphical_model.pdf}
            \label{fig:graphical model}
        }
        \subfloat[][]{
            \includegraphics[width=0.25\textwidth]{EDHMM_aux_graphical_model.pdf}
         \label{fig:aux graphical model}
        }
        \caption{a) The Explicit Duration Hidden Markov Model. The time left in the current state $x_t$ is denoted $d_t$. The observation at each point in time is denoted $y_t$. b) The EDHMM with the additional auxiliary variable $u_t$ used in the beam sampler.}
    \label{fig:graphs}    
    \end{figure}
\section{Explicit Duration Hidden Markov Model}
\label{sec:Model}

@@

The need to accommodate explicit state duration distributions in HMMs has long been recognised. Rabiner \cite{Rabiner89} details the basic approach which expands the state space to include dwell time before applying a slightly modified Baum-Welch algorithm. This approach specifies a maximum state duration, limiting practical application 
to cases with short sequences and dwell times.
This approach, generalised under the name ``segmental hidden Markov models'', includes more general transitions than those Rabiner considered, allowing the next state and duration to be conditioned on the previous state and duration \cite{Gales93}. Efficient approximate inference procedures were developed in the context of speech recognition \cite{Ostendorf96} and evolved into symmetric approaches suitable for implementation \emph{in-silico} \cite{Yu2006}.

@@

\bibliography{library}

>>>>

\documentclass[journal]{../inc/oldtran/IEEEtran}
\usepackage{amsmath,amssymb,graphicx,subfig}
\usepackage[usenames,dvipsnames]{color}
\usepackage{algorithm,algorithmic}
\newcommand{\mike}[1]{\textcolor{red}{\textsf{\emph{\textbf{\textcolor{red}{#1}}}}}}
\newcommand{\frank}[1]{\textcolor{red}{\textsf{\emph{\textbf{\textcolor{blue}{#1}}}}}}
\newcommand{\simeon}[1]{\textcolor{red}{\textsf{\emph{\textbf{\textcolor{green}{#1}}}}}}
\newcommand{\E}{\textrm{E}}

\title{Inference in Hidden Markov Models with Explicit State Duration Distributions}
\author{Michael Dewar, Chris Wiggins, Frank Wood}

@@

    \begin{figure}[t]
        \centering
        \subfloat[][]{
            \includegraphics[width=0.25\textwidth]{../pic/EDHMM_graphical_model.pdf}
            \label{fig:graphical model}
        }
        \subfloat[][]{
            \includegraphics[width=0.25\textwidth]{../pic/EDHMM_aux_graphical_model.pdf}
         \label{fig:aux graphical model}
        }
        \caption{a) The Explicit Duration Hidden Markov Model. The time left in the current state $x_t$ is denoted $d_t$. The observation at each point in time is denoted $y_t$. b) The EDHMM with the additional auxiliary variable $u_t$ used in the beam sampler.}
    \label{fig:graphs}    
    \end{figure}
\section{Explicit Duration Hidden Markov Model}
\label{sec:Model}

@@

The need to accommodate explicit state duration distributions in HMMs has long been recognised. Rabiner \cite{Rabiner89} details the basic approach which expands the state space to include dwell time before applying a slightly modified Baum-Welch algorithm. This approach specifies a maximum state duration, limiting practical application 
to cases with short sequences and dwell times.
This approach, generalised under the name ``segmental hidden Markov models'', includes more general transitions than those Rabiner considered, allowing the next state and duration to be conditioned on the previous state and duration \cite{Gales93}. Efficient approximate inference procedures were developed in the context of speech recognition \cite{Ostendorf96} and evolved into symmetric approaches suitable for practical implementation \cite{Yu2006}.

@@

\bibliography{../inc/library}

### 2011-10-27 tex/introduction.tex
small edits, mostly to do with computation complexity. One graph left...

In demonstrating our EDHMM inference technique we consider a synthetic system  in which the state-cardinality is known and finite and the distributions over state durations are of scientific interest, we focus on the finite state, unknown duration distribution case. We show that the EDHMM beam sampler performs accurate tracking whilst capturing the duration distributions over the dwell time of each state as well as the probability of transitioning between states.

>>>>

In demonstrating our EDHMM inference technique we consider a synthetic system  in which the state-cardinality is known and finite, but where each state's duration distribution is unknown. We show that the EDHMM beam sampler performs accurate tracking whilst capturing the duration distributions as well as the probability of transitioning between states.

### 2011-10-27 tex/results.tex
updated paths
small edits, mostly to do with computation complexity. One graph left...

\begin{figure}
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{experiment_2_X.pdf}
        \label{fig:exp_1_state}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{experiment_2_Y.pdf}
        \label{fig:exp_1_data}
    }
    \caption{Example a) state and b) observation sequence generated by the explicit duration HMM. Here $K$ = 3; $p(y_t|x_t=j) = \mathrm{N}(\mu_j, 1)$ with $\mu_1 = -3$, $\mu_2 = 0$, and $\mu_3 = 3$; and $p(d_t|x_t=j) = \mathrm{Poisson}(\lambda_j)$ with $\lambda_1 = 5$, $\lambda_2 = 15$, and $\lambda_3 = 20$.}
    \label{fig:experiment1_data}
\end{figure}

\begin{figure}
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{posterior_means.pdf}
        \label{fig:posterior_means}
    } 
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{posterior_rates.pdf}
        \label{fig:posterior_rates}
    }
    \caption{Samples from the posterior distributions of a) the observation distribution means and b) the duration distribution rate parameters.}
    \label{fig:experiment1_results}
\end{figure}
\subsection{Synthetic Data}

@@

\begin{figure}
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{experiment_3_Y.pdf}
        \label{fig:exp_2_data}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{experiment_3_X.pdf}
        \label{fig:exp_2_state}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{posterior_means_exp3.pdf}
        \label{fig:posterior_means_3}
    } 
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{posterior_rates_exp3.pdf}
        \label{fig:posterior_rates_3}
    }
    \caption{Results of the beam sampler applied to a system that generates states with identical observation distribution but differing durations. The observations are shown in a), and the true states are shown in b) overlaid with 20 randomly selected state traces produced by the sampler after burn-in. Samples from the posterior observation distribution mean are shown in c), and samples from the posterior duration distribution rates are shown in d).}
    \label{fig:experiment2_results}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=0.5\textwidth]{number_transitions_visited.pdf}

>>>>

\begin{figure}
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{../pic/experiment_2_X.pdf}
        \label{fig:exp_1_state}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{../pic/experiment_2_Y.pdf}
        \label{fig:exp_1_data}
    }
    \caption{Example a) state and b) observation sequence generated by the explicit duration HMM. Here $K$ = 3; $p(y_t|x_t=j) = \mathrm{N}(\mu_j, 1)$ with $\mu_1 = -3$, $\mu_2 = 0$, and $\mu_3 = 3$; and $p(d_t|x_t=j) = \mathrm{Poisson}(\lambda_j)$ with $\lambda_1 = 5$, $\lambda_2 = 15$, and $\lambda_3 = 20$.}
    \label{fig:experiment1_data}
\end{figure}

\begin{figure}
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_means.pdf}
        \label{fig:posterior_means}
    } 
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_rates.pdf}
        \label{fig:posterior_rates}
    }
    \caption{Samples from the posterior distributions of a) the observation distribution means and b) the duration distribution rate parameters using the model described in Figure \ref{fig:exp_1_data}.}
    \label{fig:experiment1_results}
\end{figure}
\subsection{Synthetic Data}

@@

\begin{figure}
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{../pic/experiment_3_Y.pdf}
        \label{fig:exp_2_data}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.5\textwidth]{../pic/experiment_3_X.pdf}
        \label{fig:exp_2_state}
    } \\
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_means_exp3.pdf}
        \label{fig:posterior_means_3}
    } 
    \subfloat[][]{
        \includegraphics[width=0.25\textwidth]{../pic/posterior_rates_exp3.pdf}
        \label{fig:posterior_rates_3}
    }
    \caption{Results of the beam sampler applied to a system that generates states with identical observation distributions but differing durations. The observations are shown in a), and the true states are shown in b) overlaid with 20 randomly selected state traces produced by the sampler. Here the observation distributions have means $\mu_1 = 0$, $\mu_2 = 0$, and $\mu_3 = 3$. Samples from the posterior observation distribution mean are shown in c), and samples from the posterior duration distribution rates are shown in d).}
    \label{fig:experiment2_results}
\end{figure}

@@

\begin{figure}
    \includegraphics[width=0.5\textwidth]{../pic/number_transitions_visited.pdf}

### 2011-10-27 tex/sampling.tex
small edits, mostly to do with computation complexity. One graph left...

Unfortunately the sum in \eqref{eqn:forward recursion} has an infinite number of terms in the usual case of  duration distributions with countably infinite support. The standard approach to EDHMM inference involves truncating these sums in order to make them tractable. 
%If all durations are truncated at a maximum $d_\mathrm{max}$, yet many states are of duration far less than $d_{\mathrm{max}}$, this necessarily leads to much unnecessary computation
% if not all states have durations near $d_\mathrm{max}$. 
A reasonable computational artifice is to truncate all durations at some maximum $d_\mathrm{max}$; however, this leads to tremendous unneccesary computation for those states of duration much less than the chosen maximum.
In general, setting allowable duration intervals $d_\mathrm{min}<d<d_\mathrm{max}$
is highly problematic; moreover, poor choices can lead to poor inference.

%The complexity of the standard approach is always $O(T(Kd_\mathrm{max})^2)$ which is highly inefficient for segments of activity that are always long. The beam sampler below is only constrained by the effective support of the duration distribution, and hence can deal with extremely long durations with no penalty. The worst-case complexity of the beam-sampler would be achieved with a uniform duration distribution with support over $[1, d_\mathrm{max}]$.

@@

A recent contribution to inference 
in 
%is 
the infinite Hidden Markov Model (iHMM) \cite{Beal2002} suggests a way around truncation \cite{vanGael2008}.  The iHMM is an HMM with a countable number of states.  Computing the forward message for a forward filtering, backward sampler for the latent states in an iHMM also requires a sum over a countable number of elements.  
% which can be dealt with using the beam sampler \cite{vanGael2008}. Here, instead of a %potentially infinite duration-space cardinality, the iHMM faces a potentially infinite state cardinality. 
%The beam sampler combines dynamic programming with slice sampling \cite{Neal2003} in order to sample from the hierarchical Dirichlet process inherent to the iHMM.
The ``beam sampling'' approach  \cite{vanGael2008}, which we can apply largely without modification, is to truncate this sum by introducing a ``slice''  (see \cite{Neal2003}) auxiliary variable $\mathcal{U} = \{u_1, u_2, \ldots,u_T\}$ at each time step.  The auxiliary variables are chosen in such a way as to automatically limit each sum in the forward pass to a finite number of terms while allowing inference to remain exact.% in the asymptotic sampling limit.

@@

The particular choice of auxiliary variable $u_t$ is important.  We follow  \cite{vanGael2008} in choosing $u_t$ to be conditionally distributed given the current and previous state and duration in the following way (see the graphical model in Figure \ref{fig:aux graphical model}):
\begin{equation}
    \label{eqn:slice}
    p(u_t | z_t, z_{t-1}) = 
    \frac
    {\mathbb{I}(0 < u_t < p(z_t | z_{t-1}))} 
    {p(z_t | z_{t-1})}.
\end{equation}
Given $\mathcal{U}$ it possible to sample the state $\mathcal{X}$ and duration $\mathcal{D}$ conditional posterior. 
%\begin{equation}
%    \hat{\alpha}_t(z_t) = p(z_t, d_t , \mathcal{Y}_1^t, \mathcal{U}_1^t)
%\end{equation}
Using notation $\mathcal{Y}_{t_1}^{t_2} = \{y_{t_1}, y_{t_1+1}, \ldots,y_{t_2}\}$  to indicate sub-ranges of a sequence, the new forward messages we compute are:
\begin{eqnarray}
   \hat{\alpha}_t(z_t) &=& 
   p(z_t, \mathcal{Y}_1^t , \mathcal{U}_1^{t-1})   \label{eqn:scaled forward} \\
   &=& 
   \sum_{z_{t-1}}
   p(z_t, z_{t-1} , \mathcal{Y}_1^t , \mathcal{U}_1^{t-1}) \nonumber \\
   &\propto& 
   \sum_{z_{t-1}}
   p(u_{t-1} | z_t, z_{t-1})
   p(z_t, z_{t-1} , \mathcal{Y}_1^t, \mathcal{U}_1^{t-1}) \nonumber \\
   &=& 
   \sum_{z_{t-1}}
   \mathbb{I}(0 < u_{t-1} < p(z_t | z_{t-1}))
   p(y_t|z_t) \hat{\alpha}_{t-1}(z_{t-1}) \nonumber.
\end{eqnarray}
The indicator function results in non-zero probabilities in the forward message for only those states $z_t$ whose likelihood given $z_{t-1}$ is greater than $u_t$. The set of $z_t$'s for which this is true will always be finite. %\mike{PROVE THIS OR ADD A SENTENCE}
%Hence \eqref{eqn:indicator} can be rewritten as a limitation of the sum over possible state transitions:
%\begin{equation}
%    \hat{\alpha}_t(z_t) = 
%    \begin{cases} 
%            p(y_t|z_t)\sum_{z_{t-1}}\hat{\alpha}_{t-1}(z_{t-1}) & \textrm{if $u_t < p(z_t | z_{t-1})$} \\
%            0 & \textrm{otherwise}
%        \end{cases}    
%    \label{eqn:scaled forward}
%\end{equation}
%which has restricted the calculation to a finite set. 
%Care must be taken to find those transitions $p(z_t | z_{t-1})$ whose likelihood is greater than the auxiliary variable $u_t$, though this is a straightforward task for unimodal distributions.

>>>>

Unfortunately the sum in \eqref{eqn:forward recursion} has an infinite number of terms in the usual case of  duration distributions with countably infinite support. The standard approach to EDHMM inference involves truncating these sums in order to make them tractable. 
%If all durations are truncated at a maximum $d_\mathrm{max}$, yet many states are of duration far less than $d_{\mathrm{max}}$, this necessarily leads to much unnecessary computation
% if not all states have durations near $d_\mathrm{max}$. 
A reasonable computational artifice is to truncate all durations at some maximum $d_\mathrm{max}$, or to consider all possible durations up to the observed length of the sequence. This leads to either tremendous unnecessary computation, or inference that simply fails if an observed duration occurs outside the allowable duration intervals $d_\mathrm{min}<d<d_\mathrm{max}$.

The complexity of the standard approach \cite{Yu10} is always $O(T(Kd_\mathrm{max})^2)$ which is highly inefficient for segments of activity that are always long. The beam sampler below is only constrained by the effective support of the duration distribution, and hence can deal with extremely long durations with no penalty. The worst-case complexity of the beam-sampler would be achieved with a uniform duration distribution with support over $[1, d_\mathrm{max}]$.

@@

A recent contribution to inference 
in 
%is 
the infinite Hidden Markov Model (iHMM) \cite{Beal2002} suggests a way around truncation \cite{vanGael2008}.  The iHMM is an HMM with a countable number of states.  Computing the forward message for a forward filtering, backward sampler for the latent states in an iHMM also requires a sum over a countable number of elements.  
% which can be dealt with using the beam sampler \cite{vanGael2008}. Here, instead of a %potentially infinite duration-space cardinality, the iHMM faces a potentially infinite state cardinality. 
%The beam sampler combines dynamic programming with slice sampling \cite{Neal2003} in order to sample from the hierarchical Dirichlet process inherent to the iHMM.
The ``beam sampling'' approach  \cite{vanGael2008}, which we can apply largely without modification, is to truncate this sum by introducing a ``slice''  (see \cite{Neal2003}) auxiliary variable $\mathcal{U} = \{u_1, u_2, \ldots,u_T\}$ at each time step.  The auxiliary variables are chosen in such a way as to automatically limit each sum in the forward pass to a finite number of terms while still allowing all possible durations.% in the asymptotic sampling limit.

@@

The particular choice of auxiliary variable $u_t$ is important.  We follow  \cite{vanGael2008} in choosing $u_t$ to be conditionally distributed given the current and previous state and duration in the following way (see the graphical model in Figure \ref{fig:aux graphical model}):
\begin{equation}
    \label{eqn:slice}
    p(u_t | z_t, z_{t-1}) = 
    \frac
    {\mathbb{I}(0 < u_t < p(z_t | z_{t-1}))} 
    {p(z_t | z_{t-1})}.
\end{equation}
Given $\mathcal{U}$ it possible to sample the state $\mathcal{X}$ and duration $\mathcal{D}$ conditional posterior. 
%\begin{equation}
%    \hat{\alpha}_t(z_t) = p(z_t, d_t , \mathcal{Y}_1^t, \mathcal{U}_1^t)
%\end{equation}
Using notation $\mathcal{Y}_{t_1}^{t_2} = \{y_{t_1}, y_{t_1+1}, \ldots,y_{t_2}\}$  to indicate sub-ranges of a sequence, the new forward messages we compute are:
\begin{eqnarray}
   \hat{\alpha}_t(z_t) &=& 
   p(z_t, \mathcal{Y}_1^t , \mathcal{U}_1^{t-1})   \label{eqn:scaled forward} \\
   &=& 
   \sum_{z_{t-1}}
   p(z_t, z_{t-1} , \mathcal{Y}_1^t , \mathcal{U}_1^{t-1}) \nonumber \\
   &\propto& 
   \sum_{z_{t-1}}
   p(u_{t-1} | z_t, z_{t-1})
   p(z_t, z_{t-1} , \mathcal{Y}_1^t, \mathcal{U}_1^{t-1}) \nonumber \\
   &=& 
   \sum_{z_{t-1}}
   \mathbb{I}(0 < u_{t-1} < p(z_t | z_{t-1}))
   p(y_t|z_t) \hat{\alpha}_{t-1}(z_{t-1}) \nonumber.
\end{eqnarray}
The indicator function results in non-zero probabilities in the forward message for only those states $z_t$ whose likelihood given $z_{t-1}$ is greater than $u_t$. The set of $z_t$s for which this is true will always be finite and smaller than the worst case $(Kd_\mathrm{max})^2$. 
%Hence \eqref{eqn:indicator} can be rewritten as a limitation of the sum over possible state transitions:
%\begin{equation}
%    \hat{\alpha}_t(z_t) = 
%    \begin{cases} 
%            p(y_t|z_t)\sum_{z_{t-1}}\hat{\alpha}_{t-1}(z_{t-1}) & \textrm{if $u_t < p(z_t | z_{t-1})$} \\
%            0 & \textrm{otherwise}
%        \end{cases}    
%    \label{eqn:scaled forward}
%\end{equation}
%which has restricted the calculation to a finite set. 
%Care must be taken to find those transitions $p(z_t | z_{t-1})$ whose likelihood is greater than the auxiliary variable $u_t$, though this is a straightforward task for unimodal distributions.
